---
title: "Graphs for exploration and publication"
output: pdf_document
layout: page
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plotting our data is one of the best ways to
quickly explore it and the various relationships
between variables.

There are three main plotting systems in R,
the [base plotting system][base], the [lattice][lattice]
package, and the [ggplot2][ggplot2] package.

[base]: http://www.statmethods.net/graphs/
[lattice]: http://www.statmethods.net/advgraphs/trellis.html
[ggplot2]: http://www.statmethods.net/advgraphs/ggplot2.html

Today we'll be learning about the ggplot2 package, because
it is the most effective for creating publication quality
graphics.

ggplot2 is built on the grammar of graphics, the idea that any plot can be
expressed from the same set of components: a **data** set, a
**coordinate system**, and a set of **geoms**--the visual representation of data
points.

The key to understanding ggplot2 is thinking about a figure in layers: just like
you might do in an image editing program like Photoshop, Illustrator, or
Inkscape.

Let's start off with an example:
```{r, echo=FALSE}
library(curl)
library(plyr)
gapminder_location<-curl(url = "https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
gapminder<-read.csv(gapminder_location)
```


```{r}
library(ggplot2)
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

So the first thing we do is call the `ggplot` function. This function lets R
know that we're creating a new plot, and any of the arguments we give the
`ggplot` function are the *global* options for the plot: they apply to all
layers on the plot.

We've passed in two arguments to `ggplot`. First, we tell `ggplot` what data we
want to show on our figure, in this example the gapminder data we read in
earlier. For the second argument we passed in the `aes` function, which
tells `ggplot` how variables in the **data** map to *aesthetic* properties of
the figure, in this case the **x** and **y** locations. Here we told `ggplot` we
want to plot the "lifeExp" column of the gapminder data frame on the x-axis, and
the "gdpPercap" column on the y-axis. Notice that we didn't need to explicitly
pass `aes` these columns (e.g. `x = gapminder[, "lifeExp"]`), this is because
`ggplot` is smart enough to know to look in the **data** for that column!

By itself, the call to `ggplot` isn't enough to draw a figure:


```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap))
```

We need to tell `ggplot` how we want to visually represent the data, which we
do by adding a new **geom** layer. In our example, we used `geom_point`, which
tells `ggplot` we want to visually represent the relationship between **x** and
**y** as a scatterplot of points:


```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

#### Challenge 1

Modify the example so that the figure visualise how life expectancy has
changed over time:

```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) + geom_point()
```

Hint: the gapminder dataset has a column called "year", which should appear
on the x-axis.

#### Challenge 2
In the previous examples and challenge we've used the `aes` function to tell
the scatterplot **geom** about the **x** and **y** locations of each point.
Another *aesthetic* property we can modify is the point *color*. Modify the
code from the previous challenge to **color** the points by the "continent"
column. What trends do you see in the data? Are they what you expected?

Sometimes you will want to save plots. The function
`pdf` creates a new pdf device. You can control the size and resolution
using the arguments to this function. You then create your plot, which goes to the pdf device. Finally, you have to make sure to turn off the pdf device.

```{r}
pdf("Life_Exp_vs_time.pdf", width=12, height=4)
ggplot(data=gapminder, aes(x=year, y=lifeExp, colour=country)) +
  geom_line()
dev.off()
```


## Layers

Using a scatterplot probably isn't the best for visualising change over time.
Instead, let's tell `ggplot` to visualise the data as a line plot:


```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line()
```

Instead of adding a `geom_point` layer, we've added a `geom_line` layer. We've
added the **by** *aesthetic*, which tells `ggplot` to draw a line for each
country.

But what if we want to visualise both lines and points on the plot? We can
simply add another layer to the plot:


```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country, color=continent)) +
  geom_line() + geom_point()
```

It's important to note that each layer is drawn on top of the previous layer. In
this example, the points have been drawn *on top of* the lines. Here's a
demonstration:


```{r}
ggplot(data = gapminder, aes(x=year, y=lifeExp, by=country)) +
  geom_line(aes(color=continent)) + geom_point()
```


In this example, the *aesthetic* mapping of **color** has been moved from the
global plot options in `ggplot` to the `geom_line` layer so it no longer applies
to the points. Now we can clearly see that the points are drawn on top of the
lines.

#### Transformations and statistics

ggplot also makes it easy to overlay statistical models over the data. To
demonstrate we'll go back to our first example:

```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```


Currently it's hard to see the relationship between the points due to some strong
outliers in GDP per capita. We can change the scale of units on the y axis using
the *scale* functions. These control the mapping between the data values and
visual values of an aesthetic.


```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10()
```

The `log10` function applied a transformation to the values of the gdpPercap
column before rendering them on the plot, so that each multiple of 10 now only
corresponds to an increase in 1 on the transformed scale, e.g. a GDP per capita
of 1,000 is now 3 on the y axis, a value of 10,000 corresponds to 4 on the y
axis and so on. This makes it easier to visualise the spread of data on the
y-axis.

We can fit a simple relationship to the data by adding another layer,
`geom_smooth`:

```{r}
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() + scale_y_log10() + geom_smooth(method="lm")
```


#### Challenge 3
Create a function that outputs two pdf files, each containing a plot of the data. 
Call the function from the main part of your script.